---
title: "NOAA-Storm-Analysis"
author: "Max Lyons"
date: "July 9, 2016"
output: html_document
---

## Synopsis
This is an analysis of the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database for the reproducible research course
on Coursera. The goal of the analysis is to determine which types of storm 
events have the greatest impact on 1) population health and 2) the economy
across the U.S.


## Data Processing
The first step in our analysis is pulling the data from the Coursera website,
storing it in a dataframe, and caching the result since it's a moderately
large dataset:
```{r, cache=TRUE}
library(dplyr)
data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(data_url,"NOAAdata.csv.bz2", method="curl")
storms <- read.csv("NOAAdata.csv.bz2",stringsAsFactors = FALSE)
```
Let's quickly check the structure of the dataset to see what fields are 
available and compare to the online information about the storm database.
```{r}
str(storms)
```
The dataset appears to have information about a few major categories: location,
time span, fatalities/injuries, and damage. Since we know that we want to 
identify the types of events that have the biggest impact on population health 
and the economy, we will focus on the fatalities/injuries and damage data
### Fatalities/Injuries Data
According to the National Weather Service Documentation, Direct fatalities and
injuries from storms are classified in the FATALITIES and INJURIES columns.
```{r}
summary(storms$FATALITIES)
summary(storms$INJURIES)
```
From basic summaries of the data, we can tell that most storms don't cause a 
single injury or fatality, as the median of both columns is 0. Let's figure out
what percentage of storms have at least one injury or fatality, and then look at
a summary of that subset of data
```{r}
sum(storms$FATALITIES[storms$FATALITIES >= 1])/nrow(storms)*100
sum(storms$INJURIES[storms$INJURIES >= 1])/nrow(storms)*100
```



## Results