---
title: "US Storms with the most harmful public health and economic consequences"
author: "Max Lyons"
date: "July 9, 2016"
output: html_document
---

## Synopsis
This is an analysis of the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database, with the goal of determining which types
of storm events have the most harmful impact on 1) population health and 
2) the economy across the U.S. The data comes from the national weather service
and includes storm events starting in 1950 up through November 2011. The count
of injuries and fatalities for each storm type was used as a measure of the 
harm to public health, and the analysis showed that both heat and excessive heat
were most harmful to publich health in terms of both injuries and fatalities. 
To measure the economic impact, the total US dollar value in damages was used,
and the results showed that hurricanes had the highest median damage value of
all storm types along with the second highest mean, which suggests that any
given hurricane has the most likelihood of having a negative impact on the 
economy.


## Data Processing
Lets load some packages that we will use throughout the analysis for cleaning, 
graphing, and presenting the data.
```{r}
library(dplyr)
library(ggplot2)
library(xtable)
```

The first step in our analysis is pulling the data from the Coursera website,
storing it in a dataframe, and caching the result since it's a moderately
large dataset:
```{r, cache=TRUE}
data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(data_url,"NOAAdata.csv.bz2", method="curl")
storms <- read.csv("NOAAdata.csv.bz2",stringsAsFactors = FALSE)
```
Let's quickly check the structure of the dataset to see what fields are 
available and compare to the online documentation about the storm database.
```{r}
str(storms)
```
The dataset appears to have information about a few major categories: location,
time span, fatalities/injuries, and damage. Since we know that we want to 
identify the types of events that have the biggest impact on population health 
and the economy, we will focus on the fatalities/injuries, damage, and storm 
event type data.
```{r}
storms_final_fields <- storms %>% 
    select(EVTYPE, FATALITIES, INJURIES, PROPDMG:CROPDMGEXP, REMARKS)
```
Let's look at each of our fields to see what kind of preprocessing we may need
to do:
### EVTYPE
```{r}
storm_types <- storms_final_fields %>% count(EVTYPE) %>% arrange(-n)
storm_types
summary(storm_types$n)
```
We can see that there are 985 different labels for storms, and looking at the
summary of the counts for each type, at least 75% of the labels have occurences
of 5 or less. We will need to go through and clean up some of the labels, but
let's see if we can reduce the total number of labels by focusing on the types
of storms that effect population health or the economy.
### Population Health factors
According to the National Weather Service Documentation, Direct fatalities and
injuries from storms are classified in the FATALITIES and INJURIES columns. We 
will use these columns without any additional processing to evaluate population
health.
### Economic factors
There are four fields that are utilized to provide information about the 
economic impact of the storms: PROPDMG, PROPDMGEXP, CROPDMG, and CROPDMGEXP.
Both PROPDMG and CROPDMG list the damage amounts for property and crops, 
respectively, and the other two columns give a unit of measure. We will need to
combine the values and unit of measure columns to compare damage across all 
storms. Let's check out the different units of measure:
```{r}
table(storms_final_fields$PROPDMGEXP)
table(storms_final_fields$CROPDMGEXP)
```
From these tables we can see units of measure for thousands (k/K), 
millions (M/m), and billions (B/b). I will also assume that (h/H) is hundreds, 
and we will have to investigate the (?,-,+) and the single number values:
```{r}
check_uom <- c('?','+','-','0','1','2','3','4','5','6','7','8')
questions <- storms_final_fields %>% filter((PROPDMGEXP %in% check_uom) | 
                                    (CROPDMGEXP %in% check_uom))
```
After looking into all the strange units of measure it appears that there isn't
any specific pattern to having a number as the unit of measure, but there are 
multiple instances in the Remarks that mentions damage of some kind. Since the 
most common unit of measure by far is thousands, I will simply replace any of 
these strange units of measure with thousands.
```{r}
storms_clean <- storms_final_fields %>% 
    mutate(PROPDMGEXP = ifelse(PROPDMGEXP %in% check_uom, 'K', 
                               toupper(PROPDMGEXP)), 
           CROPDMGEXP = ifelse(CROPDMGEXP %in% check_uom, 'K', 
                               toupper(CROPDMGEXP)))
## Replace the units of measure with numbers
clean_uom <- data.frame(old_uom = c('H','K','M','B',''), 
                        new_uom = c(100,1000,1000000,1000000000,0))
storms_clean <- storms_clean %>% left_join(clean_uom, 
                                           by = c('PROPDMGEXP' = 'old_uom'))
storms_clean <- storms_clean %>% left_join(clean_uom, 
                                           by = c('CROPDMGEXP' = 'old_uom'))
storms_clean <- storms_clean %>% mutate(TOTDMG = PROPDMG * new_uom.x + 
                                            CROPDMG * new_uom.y)
storms_final <- storms_clean %>% select(EVTYPE, FATALITIES, INJURIES, TOTDMG)
head(storms_final)
```
Now that we have clean measures of damage, injuries, and fatalities, all that
is left to do is clean the EVTYPE for analysis. Let's examine only the EVTYPEs
that have had damage, fatalities, or injuries:
```{r}
storm_types_new <- storms_final %>% 
    filter((FATALITIES > 0) | (INJURIES > 0) | (TOTDMG > 0)) %>%
    count(EVTYPE) %>% arrange(-n)
storm_types_new
summary(storm_types_new$n)
```
There are now only half as many storm categorizations, but we can make a few
clean ups to try and consolidate them even more:
```{r}
storms_final$EVTYPE_NEW <- gsub("TSTM ","THUNDERSTORM ",storms_final$EVTYPE)
storms_final$EVTYPE_NEW <- gsub("WINDS","WIND",storms_final$EVTYPE_NEW)
storms_final$EVTYPE_NEW <- toupper(storms_final$EVTYPE_NEW)
```

There are many more cleanups that could be done, but if we also filter out
all the storms that have less than 100 occurrences, we will have a much more 
clean data set.
```{r}
storm_types_new <- storms_final %>% 
    filter((FATALITIES > 0) | (INJURIES > 0) | (TOTDMG > 0)) %>%
    count(EVTYPE_NEW) %>% filter(n >= 100)
storms_final2 <- storms_final %>% 
    filter(EVTYPE_NEW %in% storm_types_new$EVTYPE_NEW)
perc_remain <- nrow(storms_final2)/nrow(storms_final)*100
```
We lost some of the original data but still have `r perc_remain`% of the
original data good for analysis.

## Results
We can now use our clean storm data to analyze which storm types have the 
largest impact on population health and the economy on average.
### Analysis of Impact on Population Health
To analyze which storm types most adversely affect population health we can 
analyze fatalities and injuries separately and combine our analysis as 
appropriate.
```{r, results="asis"}
average_fatalities <- storms_final2 %>% group_by(EVTYPE_NEW) %>% 
    summarise(n = n(), AVG_FAT = round(mean(FATALITIES),2), 
              MED_FAT = median(FATALITIES), 
              MAX_FAT = max(FATALITIES)) %>%
    arrange(-AVG_FAT) %>% top_n(10, wt = AVG_FAT)
average_fatalities_table <- xtable(average_fatalities)
print(average_fatalities_table, type = "html")
```

This data shows us that heat/excessive heat have by far the highest mean
fatalities of any storm type, and some very large maximums for any single storm.
To see if those high maximums are just driving up the mean let's look at the
percentage of occurrences that had a fatality:
```{r}
top_fatality_storms <- storms_final2 %>% 
    filter(EVTYPE_NEW %in% average_fatalities$EVTYPE_NEW) %>% 
    group_by(EVTYPE_NEW) %>%
    summarise(fatality_perc = mean(FATALITIES > 0)) %>%
    arrange(-fatality_perc)
p <- ggplot(top_fatality_storms, aes(y = fatality_perc, 
                                     x = factor(EVTYPE_NEW, 
                                    levels = top_fatality_storms$EVTYPE_NEW))) +
    geom_bar(stat='identity')
p <- p + ggtitle("Fatality % for High Average-Fatality Storm Types") +
    labs(x = "Storm Type", y = "% of Occurrences with Fatality")
p <- p + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
p
```
Rip currents have the highest percentage of fatalities, but after looking into
the documentation it looks like this number is driven up because a
rip current is only recorded if a drowning or near-drowning rescue is attempted.
Because of this we can feel comfortable stating that heat/excessive heat is the
most deadly storm type.  
Let's take a look at just general injuries to see if our results are similar:
```{r, results="asis"}
average_injuries <- storms_final2 %>% group_by(EVTYPE_NEW) %>% 
    summarise(n = n(), AVG_INJ = round(mean(INJURIES),2), 
              MED_INJ = median(INJURIES), 
              MAX_INJ = max(INJURIES)) %>%
    arrange(-AVG_INJ) %>% top_n(10, wt = AVG_INJ)
average_injuries_table <- xtable(average_injuries)
print(average_injuries_table, type = "html")
```

The table data for injuries shows heat and excessive heat as having the most 
average injuries per occurrence as well. This even lines up with our fatality
information as tornadoes come in third despite having some high outliers driving
up the mean.  

####Result: Heat & Excessive Heat are the most harmful storm types to health

### Analysis of Economic Impact
To analyze the effect on the economy of the different storm types we can 
calculate some different summary statistics by storm like mean, median, and
maximum:
```{r}
average_damage <- storms_final2 %>% group_by(EVTYPE_NEW) %>% 
    summarise(n = n(), AVG_DMG = mean(TOTDMG), MED_DMG = median(TOTDMG), 
              MAX_DMG = max(TOTDMG)) %>%
    arrange(-AVG_DMG) %>% top_n(5, wt = AVG_DMG)
```

Despite the fact that storm surges have the highest mean total damage of any
other storm type, Hurricanes come in second and have a much higher median of 
total damage, suggesting that any given hurricane has a higher likelihood of 
impacting the economy more drastically than any other storm type. 

####Result: Hurricanes are the most harmful storm type for the economy
